Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-31 analysis/paper/paper.qmd
rearranged the sequence of analysis

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period. We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

The introduction of new stone artefact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods. The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly debated their origin, the chronology of the Korean Late Paleolithic, and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

To understand the likely roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points. We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation]. We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal patterns in stemmed point uses? As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types depending on their hunting strategies and socio-environmental circumstances.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did]. After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021]. Understanding the emergence of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Palaeolithic, and relationships to adjacent regions in East Asia, such as Japan.

@@

Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. It is important to note that the TCSA metric alone cannot unambiguously determine artefact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts. Applying these sampling criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map). The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

@@

# compute TCSA
tcsa <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
 # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
  #omit very large TCSA, could be measurement error
  dplyr::filter(min_tcsa <= 257) %>%
  # group rare raw materials > 10
  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
  mutate(site_name_top = fct_lump_min(Site_name, 5))

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, site location, and artefact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did. We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data. For this study we include TCSA values from a total of 161 stemmed points from 32 assemblages unearthed from 26 sites.

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.

### Geographical Patterns

@@

########
mydata_ages_na <- mydata_ages %>%
  drop_na(age)

# count radiocarbon ages

n_sites <- length(unique(tcsa$Site_name))
min_n_points <- nrow(tcsa_site) # should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
# n_radiocarbon_ages <- NULL # we only have one age per assemblage
n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
```

To explore geographical patterns in TCSA values, we retained only sites that contains more than four stemmed points. This resulted in a sample of `r n_sites` sites and `r min_n_points` points. We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites. I THINK WE NEED SOMETHING MORE IN HERE. LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.

@@

@@

## Geographical patterns

We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than 4 stemmed points were grouped under the category of 'other'. Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values. This suggests that people made stemmed points for a variety of different purposes in Suyanggae. The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points. Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values. Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.

@@

## Geographical patterns of TCSA values

```{r}
#| label: fig-tcsa-per-sites
#| fig-cap: "Histograms of TCSA values by archaeological site."
#| fig-width:   4.5
#| fig-height:  4

 ggplot(tcsa_site,
              aes(min_tcsa, 
                  fill = n)) +
  geom_histogram() +
  theme_bw(base_size = 6) +
  labs(x = "TCSA", 
       y = "Count of artefacts") +
  facet_wrap(~label_site,
                 scales = "free_y",
                  ncol = 1) +
    scale_fill_viridis_c(name = "Number\nof artefacts") +
  annotate("rect", 
           xmin = 4,
           xmax = 18,
           ymin = 0,
           ymax = 2.5,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = 1.5,
               label = "Poisoned\narrow tips",
               hjust = 0,
               lineheight = 0.75,
           y = (0+2.5)/2,
           x = 5.5) +
      annotate("rect", 
           xmin = 17,
           xmax = 47,
           ymin = 2.5,
           ymax = 5,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned\narrow tips",
               hjust = 0,
              lineheight = 0.75,
           y = (2.5+5)/2,
           x = 28) +
      annotate("rect", 
           xmin = 40,
           xmax = 76,
           ymin = 5,
           ymax = 7.5,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           y = (5+7.5)/2,
           x = 53) +
      annotate("rect", 
           xmin = 42,
           xmax = 90,
           ymin = 7.5,
           ymax = 10,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           y = (7.5+10)/2,
           x = 70) +
     annotate("rect", 
           xmin = 80,
           xmax = 200,
           ymin = 10,
           ymax = 12.5,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "STabbing spear tips",
               size = point_function_group_label_size,
               hjust = 0,
           y = (10+12.5)/2,
           x = 160)
```

@@

# box plot for each MIS stage
  ggplot(tcsa_site_ass_mis, 
       aes(y = min_tcsa,
           x = factor(mis_stage, level = c('After', 'LGM', 'Before')))) +
   geom_boxplot() +
   theme_minimal() +
   theme(legend.position="none") +
   labs(y = "TCSA", 
        x = "MIS") +
  geom_quasirandom(alpha = 0.4,
                   size = 2.3) +
    scale_x_discrete(limits=rev) +
     annotate("rect", 
           xmin = 3.4,
           xmax = 3.5,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned\narrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = 3.4,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 3.5,
           xmax = 3.6,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned\narrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = 3.5,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 3.6,
           xmax = 3.7,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = 3.6,
           y = (76-40) /2 + 40) +
      annotate("rect", 
           xmin = 3.7,
           xmax = 3.8,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = 3.7,
           y = (90-42) /2 + 42) +
     annotate("rect", 
           xmin = 3.8,
           xmax = 3.9,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing\nspear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = 3.8,
           y = (200-80) /2 + 80 )

@@

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal patterns in stemmed point uses?

@@

@@

SPACIAL ANAlysis
@prates2022changes

>>>>

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.<!-- SP were mainly used before Last Glacial Maximum (LGM). SP was mainly located in certain latitude. why? --> We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods. The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly debated their origin, the chronology of the Korean Late Paleolithic, and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

To understand the likely roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points. We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation]. We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses? As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types depending on their hunting strategies and socio-environmental circumstances.

@@

Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did]. After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021]. Understanding the emergence of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Palaeolithic, and relationships to adjacent regions in East Asia, such as Japan.

@@

Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

@@

# compute TCSA
tcsa <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
 # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
  dplyr::filter(min_tcsa <= 250) %>% #omit very large TCSA, could be measurement error
  # group rare raw materials > 10
  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
  mutate(site_name_top = fct_lump_min(Site_name, 5))

@@

```{r}
# move dataframes from the Results section to create codes for numbers in the text

@@

#
mydata_ages_na <- mydata_ages %>%
  drop_na(age)

# excluded artifact #
tcsa_omit <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
  dplyr::filter (min_tcsa > 250)

# code for numbers in the text
total_site <- nrow(korean_archaeological_site_locations)
total_sp <- nrow(mydata)
total_ass <- nrow(mydata_ages)
n_sites <- length(unique(tcsa$Site_name))
min_n_points <- nrow(tcsa_site) 
# should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
# n_radiocarbon_ages <- NULL # we only have one age per assemblage
n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
n_omit <- nrow(tcsa_omit)
n_ass <- length(unique(tcsa$Assemblage))
n_raw <- nrow(tcsa_raw)
```

Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts. Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map). The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artefact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We distinguish the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and compare with the temporal pattern. We explore the relationship between TCSA values and the location of assemblages to explore spatial pattern. We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.

`r n_raw`

Different sizes of stone artifacts can constrain or enable different functions. Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment]. We chose weight as a proxy for overall size of the stemmed points. Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins]. Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.

@@

### Spatial Patterns

To explore spatial patterns in TCSA values, we retained only sites that contains more than four stemmed points. This resulted in a sample of `r n_sites` sites and `r min_n_points` points. We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites. I THINK WE NEED SOMETHING MORE IN HERE. LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.

@@

@@

@@

# box plot for each MIS stage
  ggplot(tcsa_site_ass_mis, 
       aes(y = min_tcsa,
           x = factor(mis_stage, level = c('After', 'LGM', 'Before')))) +
   geom_boxplot() +
   theme_minimal() +
   theme(legend.position="none") +
   labs(y = "TCSA", 
        x = "Time Period") +
  geom_quasirandom(alpha = 0.4,
                   size = 2.3) +
    scale_x_discrete(limits=rev) +
     annotate("rect", 
           xmin = 3.4,
           xmax = 3.5,
           ymin = 4,
           ymax = 18,
           fill = "green",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Poisoned\narrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = 3.4,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = 3.5,
           xmax = 3.6,
           ymin = 17,
           ymax = 47,
           fill = "blue",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Un-poisoned\narrow tips",
               hjust = 0,
               lineheight = 0.75,
           x = 3.5,
           y = (47-17) /2 + 17) +
      annotate("rect", 
           xmin = 3.6,
           xmax = 3.7,
           ymin = 40,
           ymax = 76,
           fill = "orange",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Dart tips",
               hjust = 0,
           x = 3.6,
           y = (76-40) /2 + 40) +
      annotate("rect", 
           xmin = 3.7,
           xmax = 3.8,
           ymin = 42,
           ymax = 90,
           fill = "purple",
           alpha = 0.2) +
      annotate("text", 
               size = point_function_group_label_size,
               label = "Javelin tips",
               hjust = 0,
           x = 3.7,
           y = (90-42) /2 + 42) +
     annotate("rect", 
           xmin = 3.8,
           xmax = 3.9,
           ymin = 80,
           ymax = 200,
           fill = "yellow",
           alpha = 0.2) +
      annotate("text", 
               label = "Stabbing\nspear tips",
               size = point_function_group_label_size,
               hjust = 0,
               lineheight = 0.75,
           x = 3.8,
           y = (200-80) /2 + 80 )

summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))

@@

- Unclear pattern, but mainly made before LGM
- Function is not closely related to the period based on p-value but before LGM show wide variation. 
 - some assemblages show wide variation while the other show narrow variation.

## Spatial patterns

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses?

@@

<--! In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did. -->

@@

SPATIAL ANAlysis @prates2022changes provde strong relationship between the temporal density and spatial distribution of megafauna and Fishtail projectile points, and proposed that this weapon was designed and used for megafauna hunting, contributing to their extinction by exploring changes in projectile point technology, as well as the body mass of potential megafaunal prey, and show that Fishtails were strongly related to the largest extinct megafaunal species.
